A formal approach but practical approach in train regulation

### 2025-06-18 paper/main.tex
Début relecture + macros + mise en page

\newcommand{\directions}{{\mathsf{D}}}
\newcommand{\turnouts}{{\mathsf{\theta}}}
\newcommand{\blocks}{{\mathsf{B}}}
\newcommand{\internalState}{{\mathcal{I}}}
\newcommand{\forward}{{\mathtt{f}}}
\newcommand{\backward}{{\mathtt{b}}}
\newcommand{\idle}{{\mathtt{*}}}
\newcommand{\sucblock}{{\mathtt{suc}}}

%% tt style
\newcommand{\sigred}{{\mbox{$\mathtt{r}$}}\xspace}
\newcommand{\siggreen}{{\mbox{$\mathtt{g}$}}\xspace}

@@

\newcommand{\su}{{\mbox{$\mathtt{startuntil}$}}\xspace}

@@

Final words on the N-scale model. Our circuit has 8 blocks, 5 switches. Each block is bidirectional with a signal at each end. Blocks size goes from 25cm to 70cm.  We operate  three trains in parallel. Due to the size of the model, we reduced our signals  to "red" and "green" status, the trains being too close one from the other. Furthermore, in case of a red signal, we make the hypothesis that the train will be able to stop before it reaches the end of the block. In practice, we stop the trains in the middle.

Routes are computed on the fly. The schedule is renewed every "day"\footnote{We use a time scale factor: one of our "day" corresponds to 20 minutes in real time.}  and we introduce some randomness in the process: beyond some routine daily planning, some trains are inserted/deleted, some trains departure/destination may change, and so on. We simulate also delays. Finally, in our model, we distinguish freight and  passengers routes giving priority to the latter. Routes are recomputed each time a train has finished its duty. The general motivation of this paper is to propose a lightweight scheduling solution that is adapted to the fully automatic control, that takes into account the distributed nature of the system, that is reliable and does not suffer from interlocking.

@@

\todo{ Comparaison avec les travaux antérieurs }

@@

The general principle of our model is to have N agents corresponding to the N trains and one agent corresponding to the Regulator. But first, we have to describe the infrastructure, that is  the playing field.

We suppose that the train network can be described as follows. First, we consider given a set of blocks: $\blocks = \{ 1, 2, \ldots, n\}$. Blocks are oriented: there is a $\forward$ (forward) end and a $\backward$ (backward) one. Let $\directions = \{\forward, \backward\}$ be the two directions.  Next, we suppose given a set $\turnouts = \{ n+1, \ldots, m\}$, the identifiers  of trackworks (switches, crossings, double slip switches, and so on). Trackworks are characterized by an internal state: switches can be in direct or deviate position (next denoted '\direct' and '\deviate'),  a crossing has only one internal position, a double slip switch can be straight through or crossing. In the examples, to simplify the presentation, we use only switches, but our model is not restricted to them.  Let $\internalState$ denotes the set of internal states. A \emph{network configuration} is given by a function $\sigma: \turnouts \to \internalState$. Let $\Sigma$ denotes the set of all network configurations. For two turnouts $\theta = \{9, 10\}$, the function $9 \mapsto \deviate, 10 \mapsto \direct$ states that $9$ is in deviate position while 10 is in direct one. The latter function is also described via an array  $[\deviate, \direct]$ giving states in trackwork order.

The network is then described as a triple $(\blocks, \turnouts, \sucblock)$ with $\sucblock: \blocks \times \directions \times \Sigma \to \blocks_\bot$. The successor function outputs the first block met according to the starting position, the direction and the network configuration.  The value $\sucblock(pid, d, \sigma) = \bot$ means that one reaches a dead-end. From an abstract point of view, the topology of the network reduces to that data.

On the tracks, we suppose there are sensors to detect the presence of trains. More precisely, we suppose that---by means of the sensors---one is able to compute at any time the position of each train. A \emph{position event} is when a train move from one block to an other one.

Finally, aside of the tracks, at each end of the blocks, there is a signal whose state is \sigred (red) or \siggreen (green). In other words, the status of signals is described by a function $F: \blocks \times \directions \to \{ \sigred, \siggreen\}$. Given some signal status $F$, a block $pid$ and a direction $d$, $F(pid,d) = \sigred$ means a train in block $pid$ cannot leave it in direction $d$. The signal status is entirely determined by the network configuration and the locking status due to the regulator. We will be back to it later.  Since we do not model train speed, we do not distinguish signals with speed limits from full-speed ones.

Now, we come to the train model. Each train has an identifier. Without loss of generality, the identifiers form a set  $\trains = \{1, \ldots, k\}$. Each train has a position on the network. The position is the index of the block it stays in or the last block it left. Next, a train has a direction in $\{\forward, \backward, \idle\}$ with $\idle$ meaning idle status.  The direction does not mean the train is running---it can be blocked by some red signal---but that its next move will be in that direction. Finally, a train has orders. Orders are given as a sequence of orders following the grammar:
$$  seq ::= \su(d,pid) $$
with $d \in \directions$ and $pid \in \blocks$. An order $\su(d, pid)$ states that the train direction must be set to $d$ and that the train runs until it reaches $pid$.  The sequence may be empty.

To sum up, trains can be viewed as 4-tuples: $t = (tid, pid, d, P)$ with $tid$ the train's identifier, $pid$ its position, $d$ its direction and $P$ the sequence of orders. The description of all trains is then a sequence $\Gamma = (t_1, \ldots, t_k)$ with $t_1$ the 4-tuple corresponding to train 1, up to $t_k$ that is the 4-tuple corresponding to train $k$.

To help the reader, by convention, we use $pid$ for blocks, $sid$ for trackworks identifiers (switch id) and $tid$ for train identifiers.

The regulator is in charge of the execution of the planning. For that sake, we use tokens. For each critical resource (either a block or a switch), we associate a token level with each train involved.

More technically, the regulator is characterized by 3 fields. First, it has a general program, next mentioned as $E$. The program is the plan of the regulator. The regulator will follow its rules. A program is a sequence of statements, each statement being composed of a guard and some actions. The guard has the shape: $ev(tid, n)$ meaning that train $tid$ just entered in the $n$-th block of its route. In case such an event occurs, the associated actions are performed. Here is the grammar of the regulator program:
%All statements are guarded by a condition $ev(tid,n)$ stating that train $tid$ reached the $n$-th step of its route. The atoms update the regulator fields. 
%A program is given as a sequence of statements. Those follow the grammar:

\begin{grammar}[Regulator's rules][h][regulateur]
    \firstcasesubtil{stmt}{ev(tid, n) : (atom)*}{}
    \firstcasesubtil{atom}{\incr(pid)}{}
    \otherform{att(pid, v)}{}
    \otherform{\turn(sid, s)}{}
    \otherform{auth}{}
 \end{grammar}
with $tid \in \trains$, $n \in  \Nat$, $pid \in \blocks$, $sid \in \turnouts$, $s\in \internalState$.  
The token field of the regulator, denoted $T$, is a function  $T: \blocks \to \Nat$ associating a token level to each block.  The atom $\incr(pid)$ updates the token $pid$:  $T \cdot \incr(pid)  \triangleq T[pid \leftarrow T(pid)+1]$.

@@

>>>>

\newcommand{\directions}{\dirFmt{\mathsf{D}}}
\newcommand{\turnouts}{{\mathsf{\theta}}}
\newcommand{\blocks}{{\posFmt{\mathsf{B}}}}
\newcommand{\internalState}{{\mathcal{I}}}

% Je suis assez d'avis de distinguer la direction des blocks et des trains
\newcommand{\forward}{\ensuremath{\mathtt{f}}}
\newcommand{\backward}{\ensuremath{\mathtt{b}}}
\newcommand{\sucblock}{{\mathtt{suc}}}
\newcommand{\networkConf}{\ensuremath{\Sigma}}

%% Train directions
\definecolor{dircolor}{rgb}{1.0, 0.13, 0.32}
\newcommand{\dirFmt}[1]{{\color{dircolor}{#1}}}
\newcommand{\dirForward}{\ensuremath{\dirFmt{\forward}}\xspace}
\newcommand{\dirBackward}{\ensuremath{\dirFmt{\backward}}\xspace}
\newcommand{\dirStop}{\ensuremath{\dirFmt{\star}}\xspace}

%% Positions
\definecolor{poscolor}{rgb}{0.6, 0.45, 0.32}
\newcommand{\posFmt}[1]{{\color{poscolor}{#1}}}

%% Trains
\newcommand{\trainTuple}[4]{\langle #1, \posFmt{#2}, \dirFmt{#3}, #4 \rangle}
\newcommand{\trainSeq}{\Gamma}

%% Regulator's Program
\newcommand{\evGuard}[2]{\ensuremath{\mathtt{ev}(#1, #2)}}

%% tt style
\newcommand{\sigred}{{\mbox{${\color{red!50!black}\mathtt{r}}$}}\xspace}
\newcommand{\siggreen}{{\mbox{${\color{green!50!black}\mathtt{g}}$}}\xspace}

@@

\newcommand{\su}[2]{{\mbox{$\mathtt{startuntil}(\dirFmt{#1}, \posFmt{#2})$}}\xspace}

@@

\paragraph{Final words on the N-scale model.} Our circuit has 8 blocks, 5 switches. Each block is bidirectional with a signal at each end. Blocks size goes from 25cm to 70cm.  We operate  three trains in parallel. Due to the size of the model, we reduced our signals  to "red" and "green" status, the trains being too close one from the other. Furthermore, in case of a red signal, we make the hypothesis that the train will be able to stop before it reaches the end of the block. In practice, we stop the trains in the middle.

\paragraph{Routing trains.}
\todo{Ajouter: tout est automatique}
Routes are computed on the fly. The schedule is renewed every "day"\footnote{We use a time scale factor: one of our "day" corresponds to 20 minutes in real time.}  and we introduce some randomness in the process: beyond some routine daily planning, some trains are inserted/deleted, some trains departure/destination may change, and so on. We simulate also delays. Finally, in our model, we distinguish freight and  passengers routes giving priority to the latter. Routes are recomputed each time a train has finished its duty. The general motivation of this paper is to propose a lightweight scheduling solution that is adapted to the fully automatic control, that takes into account the distributed nature of the system, that is reliable and does not suffer from interlocking.

@@

\todo{ Comparaison avec les travaux antérieurs / lit review}

@@

In this section, we present the foundations of our formal model, explaining the basic building blocks that we need in order to specify and verify train routing, itself presented in the following section.

The general principle of our model is to have $N$ agents corresponding to the $N$ trains and one agent corresponding to the Regulator. But first, we have to describe the infrastructure, that is  the playing field. \todo{Déplacer ?}

\subsection{Infrastructure}

\paragraph{The network.}

We suppose that the train network can be described as follows. First, we consider given a set of blocks: $\blocks = \{ 1, 2, \ldots, n\}$. Blocks are oriented: there is a $\forward$ end and a $\backward$ one. Let $\directions = \{\forward, \backward\}$ be the two directions.  Next, we suppose given a set $\turnouts = \{ n+1, \ldots, m\}$, the identifiers  of trackworks (switches, crossings, double slip switches, and so on\todo{Est-ce que ces élements sont dans le modèle de Lucas, en particulier les crossings et double slip switches ?}). Trackworks are characterized by an internal state: switches can be in direct or deviate position (next denoted '\direct' and '\deviate'), a crossing has only one internal position, a double slip switch can be straight through or crossing. In the examples, to simplify the presentation, we use only switches, but our model is not restricted to them.  Let $\internalState$ denotes the set of internal states. A \emph{network configuration} is given by a function $\sigma: \turnouts \to \internalState$. Let $\networkConf$ denotes the set of all network configurations. For two turnouts $\theta = \{9, 10\}$, the function $9 \mapsto \deviate, 10 \mapsto \direct$ states that $9$ is in deviate position while 10 is in direct one. The latter function is also described via an array  $[\deviate, \direct]$ giving states in trackwork order.

The network is then described as a triple $(\blocks, \turnouts, \sucblock)$ where $\sucblock$ is a partial function indicating the following block given a starting position, a direction and a network configuration: $\sucblock: \blocks\times \directions \times \networkConf \to \blocks$. We note $\sucblock(pid, d, \sigma) = \bot$ when it is undefined. From an abstract point of view, the topology of the network reduces to that data.

\paragraph{Signals.}

At each end of the blocks, there is a signal whose state is \sigred (red) or \siggreen (green). In other words, the status of signals is described by a function $F: \blocks \times \directions \to \{ \sigred, \siggreen\}$. Given some signal status $F$, a block $\posFmt{pos}$ and a direction $dir$, $F(pos,dir) = \sigred$ means the signal guarding the exit of block $\posFmt{pos}$ in direction $\dirFmt{dir}$ is red, and we expect trains not leaving that block going in this direction\footnote{Our semantics, presented below, ensure such behaviour.}\todo{ref}. The signal status is entirely determined by the network configuration and the locking status due to the regulator\todo{Le régulateur n'est pas encore introduit, je propose de déplacer ça plus bas.}.

\subsection{Agents}
\paragraph{Trains.}
We now present the train model.
Each train is characterised by an identifier, a position, a direction, and a list of orders.
\begin{itemize}
	\item Identifiers form a set  $\trains = \{1, \ldots, k\}$.
	\item The position is the index of the block currently occupied by the train.
	\item The direction direction in $\directions = \{\dirBackward, \dirForward, \dirStop\}$ --- where \dirBackward (resp.\ \dirForward) indicates the trains is ordered to go towards the \backward (resp.\ \forward) end of the block, while \dirStop indicates the train has no direction order at the moment. The direction represent the current move order, but does not mean the train is actually moving: it can be blocked at a signal.
	\item Orders are given as a (possibly empty) sequence of orders $\su{dir}{pos}$ ($\dirFmt{dir} \in \directions$, $\posFmt{pos} \in \blocks$). Such order states that the train direction is set to $\dirFmt{dir}$ and that the train runs until it reaches $pos$.
\end{itemize}

Then, each train is a quadruple $t = \trainTuple{id}{pos}{dir}{P}$ with $id$ the train's identifier, $\posFmt{pos}$ its position, $\dirFmt{dir}$ its direction and $P$ the sequence of orders. Trains are collected in a sequence $\trainSeq = (t_1, \ldots, t_k)$ where each $t_i$ being the tuple of train $i$.

\paragraph{The Regulator.}
The regulator is in charge of the execution of the routing.
It is characterised by a routing program, a token list, a waiting list, an history, and a switch control list.

\begin{itemize}
	\item The routing program specifies how the regulator should react to events (i.e. when a train enters a block). A program is a sequence of rules\todo{Je pense qu'il est plus clair d'appeler ça des règles que des statements}. Each rule begins by a guard \evGuard{tid}{n}, specifying that the content of the rule is triggered when the train $tid$ enters its $n$-th block.
\end{itemize}

\paragraph{The Regulator\todo{Version Guillaume}. }
More technically, the regulator is characterized by 3 fields. First, it has a general program, next mentioned as $E$. The program is the plan of the regulator. The regulator will follow its rules. A program is a sequence of statements, each statement being composed of a guard and some actions. The guard has the shape: $ev(tid, n)$ meaning that train $tid$ just entered in the $n$-th block of its route. In case such an event occurs, the associated actions are performed. Here is the grammar of the regulator program:
%All statements are guarded by a condition $ev(tid,n)$ stating that train $tid$ reached the $n$-th step of its route. The atoms update the regulator fields. 
%A program is given as a sequence of statements. Those follow the grammar:

\begin{grammar}[Regulator's rules][h][regulateur]
    \firstcasesubtil{stmt}{\evGuard{tid}{n} : \nonterm{A}^\star}{}
    \firstcase{A}{\incr(pid)}{}
    \otherform{att(pid, v)}{}
    \otherform{\turn(sid, s)}{}
    \otherform{auth}{}
 \end{grammar}
with $tid \in \trains$, $n \in  \Nat$, $pid \in \blocks$, $sid \in \turnouts$, $s\in \internalState$.  
The token field of the regulator, denoted $T$, is a function  $T: \blocks \to \Nat$ associating a token level to each block.  The atom $\incr(pid)$ updates the token $pid$:  $T \cdot \incr(pid)  \triangleq T[pid \leftarrow T(pid)+1]$.

@@

\paragraph{Network hypothesis, events, and conventions.}
On the tracks, we suppose there are sensors to detect the presence of trains. More precisely, we suppose that ---by means of the sensors--- the position of every train is known at any time. A \emph{position event} is when a train move from one block to an other one.

Train speed are not modeled, and we consider trains can always stop within their current block if they have to (e.g. if there outgoing signal of a block is \sigred, we assume the train can stop in time).

To help the reader, in the sequel, by convention, we use $pid$ for blocks, $sid$ for trackworks identifiers (switch id) and $tid$ for train identifiers.
